\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-09 submission/manuscript.tex
Additional discussion.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the surrounding seawater were also processed.
The number of reads per sample ranged between 8,407 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,726 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,407
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,007
different OTUs that ranged from 366 to 1,998 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,688.5 ± 125.6 OTUs{]} and noninvaded {[}Number of OTUs, 1,744.8 ±
150.8 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,062.8 ± 209.6 OTUs) and lower values for picoplankton
communities in the surrounding seawater (Number of OTUs, 528.1 ± 135.8
OTUs) (\autoref{supp-calculators}). Seasonal changes did not show such
large dissimilarities. Seawater communities richness was stable during
the studied period with the exception of one sampling point in December
2017 when larger values were observed. \emph{Cymodocea nodosa}
communities showed a slow increase towards the end of the study, while
\emph{Caulerpa cylindracea} (invaded and noninvaded) communities were
characterized by slightly larger values in Spring and Summer in
comparison to Autumn and Winter (\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were found
at every time point. These OTUs made also a high proportion of total
sequences (39.1 -- 51.3 \si{\percent}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of seawater
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.64, \emph{p} \textless{} 0.001).
Epiphytic microbial communities associated with \emph{Cymodocea nodosa}
showed a similar pattern (ANOSIM, R = 0.56, \emph{p} \textless{} 0.01),
while communities from the surfaces of \emph{Caulerpa cylindracea}
indicated a non so strongly supported, as in previous cases, separation
between Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31,
\emph{p} \textless{} 0.05) (\autoref{pcoa}).

@@

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{Caulerpa cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{Cymodocea nodosa} communities were consisted of
only 0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{Caulerpa cylindracea} communities their proportion was
25.7 ± 11.3 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). In addition, \emph{Caulerpa cylindracea} associated
communities were characterized by higer proportions in Winter and Summer
(invaded, 30.9 ± 12.4 \si{\percent}; noninvaded, 26.9 ± 3.4
\si{\percent}) in comparison to Autumn and Spring (invaded, 22.8 ± 10.3
\si{\percent}; noninvaded, 21.9 ± 3.8 \si{\percent}). The community was
mainly consisted of no relative \emph{Desulfobacteraceae},
\emph{Desulfatitalea}, no relative \emph{Desulfobulbaceae},
\emph{Desulfobulbus}, no relative \emph{Desulfocapsaceae},
\emph{Desulfopila}, \emph{Desulforhopalus}, \emph{Desulfotalea},
SEEP-SRB4 and uncultured \emph{Desulfocapsaceae} (\autoref{desulfo}).

@@

Since the colonization of macrophyte surfaces is performed from a pool
of seawater prokaryotic cells it is interesting to see to which extent
do these two communities differ. We observed a strong differentiation
between the surface attached and seawater communities at the level of
OTUs that is in agreement with most published studies (Burke \emph{et
al.}, 2011b; Michelou \emph{et al.}, 2013; Roth-Schulze \emph{et al.},
2016; Ugarelli \emph{et al.}, 2019). These data indicate that marine
macrophytes are selecting, from a pool of seawater microbial taxa, the
one that can colonize and proliferate on their surfaces (Salaün \emph{et
al.}, 2012; Michelou \emph{et al.}, 2013). In contrast to these findings
Fahimipour \emph{et al.} (2017) found, in a global study of
\emph{Zostera marina}, similarities between leaves and seawater samples.
Discrepancies between our data and this study could be explained by
differences in studied seagrass species, methodological variations or
biogeographic trends as Fahimipour \emph{et al.} (2017) were analyzing
samples from different locations throughout the northern hemisphere. It
is possible that ambient seawater and leaves communities from the same
location are differing but are still more similar to each other when
compared to other sampling locations. Indeed, it was found that
prokaryotic communities vary substantially between different sampling
sites (Bengtsson \emph{et al.}, 2017). When the taxonomic composition at
high ranks was analyzed no such strong differentiation was noticed.
Phyla and classes such as: \emph{Actinobacteriota}, \emph{Bacteroidota},
\emph{Cyanobacteria}, \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria} and \emph{Verrucomicrobiota}, were described
that is in agreement with previously reported data (Burke \emph{et al.},
2011b; Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013). In
contrast, when low taxonomic ranks were analyzed (i.g. family and genus)
a strong differentiation was observed. A similar differentiation at
lower taxonomic ranks was described for other species of macrophytes
(Egan \emph{et al.}, 2013; Michelou \emph{et al.}, 2013; Ugarelli
\emph{et al.}, 2019).

@@

Few studies describing microbial epiphytic communities were encompassing
seasonal shifts (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011; Michelou \emph{et al.}, 2013). Seasonal changes in the epiphytic
community were substantial as could be observed in the proportion of
OTUs that could be find at every sampling time (\(\leqslant\) 1.0
\si{\percent}). Interestingly, these OTUs were accounting for a high
proportion of sequences (\(\leqslant\) 51.3 \si{\percent}). A very
similar proportion of persistent OTUs and their sequence contribution
was reported in high frequency studies describing seasonal picoplankton
changes (Gilbert \emph{et al.}, 2009, 2012). In comparison to the
seawater community, a lower degree of seasonal shifts was observed for
the surface associated communities. It seems, that microniches on the
surfaces of macrophytes are providing more stable conditions in
comparison to the seawater. At the level of OTUs seasonal changes of
\emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{Cymodocea nodosa} was characterized by a Spring community during
maximum seagrass proliferation, a Summer community during a biomass
maximum and a Autumn/Winter community during a biomass decay. In
contrast, \emph{Caulerpa cylindracea} started to proliferate in late
Spring and was characterized only by a Summer community during maximal
biomass increase and by a Autumn/Winter/Spring community when the
biomass were at the peak and the settlement started to subsequently
decay. Similar seasonal changes in the epiphytic community was described
also for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et
al.}, 2011). The higher temporal stability of \emph{Caulerpa
cylindracea} surface communities in comparison to \emph{Cymodocea
nodosa} could be also observed in the higher proportion of shared
communities between two consecutive sampling points.

@@

@@

@@

@@

@@

@@

@@

{\centering \includegraphics[width=1\linewidth]{../results/figures/pcoa_figure}

>>>>

\usepackage{txfonts}

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the ambient seawater were also processed for
comparison. The number of reads per sample ranged between 8,407 and
77,465 sequences (\autoref{supp-nseq_notus}). Even when the highest
sequencing effort was applied the rarefaction curves did not level off
that is a common observation in high-throughput 16S rRNA amplicon
sequencing approaches (\autoref{supp-rarefaction}). Following quality
curation and exclusion of sequences mentioned before reads were
clustered into 28,726 different OTUs at a similarity level of 97
\si{\percent}. Reads numbers were normalized to the minimum number of
sequences, 8,407 (\autoref{supp-nseq_notus}), through rarefaction
resulting in 17,007 different OTUs that ranged from 366 to 1,998 OTUs
per sample (\autoref{supp-calculators}). To determine seasonal changes
of richness and diversity the Observed Number of OTUs, Chao1, ACE,
Exponential Shannon (Jost, 2006) and Inverse Simpson were calculated
after normalization through rarefaction. Generally, richness estimators
and diversity indices showed similar trends. On average, higher values
were found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,688.5 ± 125.6 OTUs{]} and noninvaded {[}Number of OTUs, 1,744.8 ±
150.8 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,062.8 ± 209.6 OTUs) and lower values for picoplankton
communities in the ambient seawater (Number of OTUs, 528.1 ± 135.8 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{Cymodocea nodosa} communities showed a slow
increase towards the end of the study, while \emph{Caulerpa cylindracea}
(invaded and noninvaded) communities were characterized by slightly
larger values in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were found
at every time point. These OTUs made also a high proportion of total
sequences (39.1 -- 51.3 \si{\percent}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of
\emph{Cymodocea nodosa} associated communities indicated a separation
between Spring, Summer and Autumn/Winter samples (ANOSIM, R = 0.56,
\emph{p} \textless{} 0.01), while communities from the surfaces of
\emph{Caulerpa cylindracea} indicated a non so strongly supported, as in
previous cases, separation between Summer and Autumn/Winter/Spring
samples (ANOSIM, R = 0.31, \emph{p} \textless{} 0.05) (\autoref{pcoa}).

@@

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{Caulerpa cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{Cymodocea nodosa} communities were consisted of
only 0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{Caulerpa cylindracea} communities their proportion was
25.7 ± 11.3 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Since the colonization of macrophyte surfaces is performed from a pool
of seawater prokaryotic cells it is interesting to see to which extent
do these two communities differ. We observed a strong differentiation
between the surface attached and seawater communities at the level of
OTUs that is in agreement with most published studies (Burke \emph{et
al.}, 2011b; Michelou \emph{et al.}, 2013; Roth-Schulze \emph{et al.},
2016; Crump \emph{et al.}, 2018; Ugarelli \emph{et al.}, 2019). These
data indicate that marine macrophytes are selecting, from a pool of
seawater microbial taxa, the one that can colonize and proliferate on
their surfaces (Salaün \emph{et al.}, 2012; Michelou \emph{et al.},
2013). In contrast to these findings Fahimipour \emph{et al.} (2017)
found, in a global study of \emph{Zostera marina}, similarities between
leaves and seawater samples. Discrepancies between our data and this
study could be explained by differences in studied seagrass species,
methodological variations or biogeographic trends as Fahimipour \emph{et
al.} (2017) were analyzing samples from different locations throughout
the northern hemisphere. It is possible that ambient seawater and leaves
communities from the same location are differing but are still more
similar to each other when compared to other sampling locations. Indeed,
it was found that prokaryotic communities vary substantially between
different sampling sites (Bengtsson \emph{et al.}, 2017). When the
taxonomic composition at high ranks was analyzed no such strong
differentiation was noticed. Phyla and classes such as:
\emph{Actinobacteriota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria} and
\emph{Verrucomicrobiota}, were described that is in agreement with
previously reported data (Burke \emph{et al.}, 2011b; Egan \emph{et
al.}, 2013; Michelou \emph{et al.}, 2013). In contrast, when low
taxonomic ranks were analyzed (i.g. family and genus) a strong
differentiation was observed. A similar differentiation at lower
taxonomic ranks was described for other species of macrophytes (Egan
\emph{et al.}, 2013; Michelou \emph{et al.}, 2013; Ugarelli \emph{et
al.}, 2019).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 51.3
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea}
associated communities were identified that could be linked to the
growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{Cymodocea nodosa} was characterized by a Spring
community during maximum seagrass proliferation, a Summer community
during a biomass maximum and a Autumn/Winter community during a biomass
decay. In contrast, \emph{Caulerpa cylindracea} started to proliferate
in late Spring and was characterized only by a Summer community during
maximal biomass increase and by a Autumn/Winter/Spring community when
the biomass were at the peak and the settlement started to subsequently
decay. Similar seasonal changes in the epiphytic community was described
also for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et
al.}, 2011). Higher temporal stability of \emph{Caulerpa cylindracea}
surface communities in comparison to \emph{Cymodocea nodosa} were also
observed in the higher proportion of shared communities between two
consecutive sampling points.

Seasonal patterns of high rank epiphytic taxa did not show strong
changes. Only \emph{Cyanobateria} were characterized by strong seasonal
differences. Cyanobacterial sequences were more pronounced in the
November and December 2017 in comparison to Spring and Summer.
Interestingly, in these high proportion months the majority of
cyanobacterial sequences were classified as \emph{Pleurocapsa}, a group
known to colonized different living and nonliving surfaces (Burns
\emph{et al.}, 2004; Longford \emph{et al.}, 2007; Mobberley \emph{et
al.}, 2012; Reisser \emph{et al.}, 2014). It is possible than during
periods of low metabolic activity of the i.g. seagrass there is no so
active interaction and selection of the epiphytic community, so the
seagrass leaves are becoming a suitable surface for nonspecific
colonizers (Zavodnik \emph{et al.}, 1998). \emph{Pleurocapsa} was
replaced in Spring ans Summer by \emph{Acrophormium},
\emph{Phormidesmis} and no relative \emph{Nodosilineaceae}. A study of
coastal microbial mats found also higher proportion of
\emph{Nodosilineaceae} sequences in Summer, while \emph{Phormidesmis}
sequences were at their peak in Autumn (Cardoso \emph{et al.}, 2019).
Other high rank taxa did not showed strong successional patterns. In
each analyzed group low rank taxa

@@

\leavevmode\hypertarget{ref-Burns2004}{}%
Burns, B.P., Goh, F., Allen, M., and Neilan, B.A. (2004) Microbial
diversity of extant stromatolites in the hypersaline marine environment
of Shark Bay, Australia. \emph{Environmental Microbiology} \textbf{6}:
1096--1101.

@@

\leavevmode\hypertarget{ref-Cardoso2019}{}%
Cardoso, D.C., Cretoiu, M.S., Stal, L.J., and Bolhuis, H. (2019)
Seasonal development of a coastal microbial mat. \emph{Scientific
Reports} \textbf{9}: 1--14.

@@

\leavevmode\hypertarget{ref-Crump2018}{}%
Crump, B.C., Wojahn, J.M., Tomas, F., and Mueller, R.S. (2018)
Metatranscriptomics and amplicon sequencing reveal mutualisms in
seagrass microbiomes. \emph{Frontiers in Microbiology} \textbf{9}:

@@

\leavevmode\hypertarget{ref-Longford2007}{}%
Longford, S., Tujula, N., Crocetti, G., Holmes, A., Holmström, C.,
Kjelleberg, S., et al. (2007) Comparisons of diversity of bacterial
communities associated with three sessile marine eukaryotes.
\emph{Aquatic Microbial Ecology} \textbf{48}: 217--229.

@@

\leavevmode\hypertarget{ref-Mobberley2012}{}%
Mobberley, J.M., Ortega, M.C., and Foster, J.S. (2012) Comparative
microbial diversity analyses of modern marine thrombolitic mats by
barcoded pyrosequencing. \emph{Environmental Microbiology} \textbf{14}:
82--100.

@@

\leavevmode\hypertarget{ref-Reisser2014}{}%
Reisser, J., Shaw, J., Hallegraeff, G., Proietti, M., Barnes, D.K.A.,
Thums, M., et al. (2014) Millimeter-sized marine plastics: a new pelagic
habitat for microorganisms and invertebrates. \emph{PloS one}
\textbf{9}: e100289.

@@

{\centering \includegraphics[width=0.55\linewidth]{../results/figures/pcoa_figure}
